---
title: "Database Design Analysis"
author: "Michael Rose"
date: "November 26, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts=list(width.cutoff=60),tidy=TRUE)
options(width = 60)
library(odbc)
library(DBI)
```


```{r, echo=FALSE}
# rmarkdown::render("final_proj.Rmd", params = "ask") 
sql_con <- dbConnect(odbc(),
                 Driver = "ODBC Driver 17 for SQL Server",
                 Server = "localhost",
                 Database = "Northwind_SPP",
                 UID = "SA",
                 PWD = rstudioapi::askForPassword("Database password"),
                 Port = 1433)
```

# Introduction 

This is an in-depth analysis of a production size database using the **MS SQL Server** dialect of SQL. 

The goal is to complete a series of queries that make use of a wide range of the functionality built within sql. 

For the sake of readability, I have limited the output to the top 25 rows. If a query has more than 25 rows, it will be noted above the output.

If you wish to see the full output, in the code there is code chunk options. Remove max.print = 25 for full output.

# SQL Analysis

## 0.1 | Tables

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT TABLE_NAME
FROM INFORMATION_SCHEMA.TABLES 
WHERE TABLE_TYPE = 'BASE TABLE'
```

## 0.2 | Table Information

```{sql, connection = sql_con, output_var = 'ex1'}
SELECT TABLE_NAME,
       COLUMN_NAME,
       DATA_TYPE
FROM   INFORMATION_SCHEMA.COLUMNS
ORDER BY TABLE_NAME
```

# Easy Queries

Functions Used: 

* Like
* Wildcard %
* In
* Date
* Order By
* Convert
* Concat
* Count
* Mutate
* Min
* Distinct
* Join

## 1 | Which Shippers Do We Have?

Return all the fields from all the shippers.

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT * 
FROM Shippers
```

## 2 | Specific Fields from Categories 

Get the categoryname and description from categories. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT CategoryName, Description 
FROM Categories
```

## 3 | Sales Representatives 

Get first name, last name, hire date of all employees with the title of sales representative 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT FirstName, LastName, HireDate 
FROM Employees 
WHERE Title = 'Sales Representative'
```

## 4 | Sales Representatives in the United States 

Constrain to employees in the USA 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT FirstName, LastName, HireDate 
FROM Employees 
WHERE Title = 'Sales Representative' 
AND Country = 'USA'
```

## 5 | Orders Placed by Specific EmployeeID 

Show all orders places by a specific employee. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT OrderID, OrderDate 
FROM Orders 
WHERE EmployeeID = 5
```

## 6 | Suppliers and Contact Titles 

In the suppliers table, we want the supplier ID, contact name and contact title for those suppliers whose contact title is not marketing manager 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT SupplierID, ContactName, ContactTitle 
FROM Suppliers 
WHERE NOT ContactTitle = 'Marketing Manager'
```

## 7 | Products with "Queso" in Product Name

We want the product ID and product name for those products where the product name includes the string Queso 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT ProductID, ProductName 
FROM Products 
WHERE ProductName 
LIKE '%queso%'
```

## 8 | Orders Shipping to France or Belgium 

We want the order id, customer id, and ship country for the orders where the ship country is either France or Belgium.
The following table shows 25 / 96 rows.

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT OrderID, CustomerID, ShipCountry 
FROM Orders 
WHERE ShipCountry 
IN ('France', 'Belgium')
```

## 9 | Orders Shipping to any Country in Latin America

We want to get all the orders from any latin american country | We don't have a list of the Latin American countries in a table, so we will make it.

This table outputs 25 / 173 rows. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT OrderID, CustomerID, ShipCountry 
FROM Orders 
WHERE ShipCountry 
IN ('Brazil', 'Mexico', 'Argentina', 'Venezuela')
```

## 10 | Employees in Order of Age 

For all the employees, show first, last, title and birth date in order of birth date with oldest employees first

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT FirstName, LastName, Title, BirthDate 
FROM Employees 
ORDER BY BirthDate ASC
```

## 11 | Showing Only the Date with a DateTime Field 

In the above query, suppose we only want the date portion of the birth date field. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT FirstName, LastName, Title, BDate = convert(DATE, BirthDate) 
FROM Employees 
ORDER BY BDate ASC
```

## 12 | Employees Full Name
 
show first, last and firstlast names

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT FirstName, LastName, FullName = CONCAT(FirstName, ' ', LastName) 
FROM Employees
```

## 13 | Order Details Amount per Line Item 

In the order details table we have the fields unit price and quantity. We want a new field, total price that multiplies these two together. We also want order id, product id, unit price, and quantity ordered by order id and product id.

This table outputs 25 / 2155 rows. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT OrderID, ProductID, UnitPrice, Quantity, TotalPrice = UnitPrice * Quantity 
FROM OrderDetails 
ORDER BY OrderID, ProductID
```

## 14 | How Many Customers? 

How many customers are there in the customers table? 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT TotalCustomers = COUNT(*) 
FROM Customers
```

## 15 | Date of First Order

Return the date of the first order 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT FirstOrder = MIN(OrderDate) 
FROM Orders
```

## 16 | Countries with Customers 

We want a list of countries where Northwind has customers 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT DISTINCT(Country) 
FROM Customers
```

## 17 | Contact Titles for Customers 

Return a list of all the different values in the customers table for contact titles. Also include a count for each contact title. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT ContactTitle, TotalContactTitle = COUNT(*) 
FROM Customers 
GROUP BY ContactTitle 
ORDER BY TotalContactTitle DESC
```

## 18 | Products with Associated Supplier Names 

For each product, we want the associated supplier. Show the product id, product name, and company name of the supplier and sort by product id. 

This table outputs 25 / 77 rows. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT ProductID, ProductName, Supplier = CompanyName 
FROM Products 
JOIN Suppliers ON Products.SupplierID = Suppliers.SupplierID
```

## 19 | Orders and the Shipper Used 

We'd like to show a list of the orders that were made, including the shipper that was used. We want order id, order date (date only), company name of the shipper, and to sort by order id with order id < 10270. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT OrderID, OrderDate = CONVERT(DATE, ORDERDATE), Shipper = CompanyName 
FROM Orders JOIN Shippers ON Orders.ShipVia = Shippers.ShipperID 
WHERE OrderID < 10270 ORDER BY OrderID
```

# Intermediate Problems 

Functions Used: 

* Exists 
* DateAdd
* Nested Select
* Multiple Join
* Join
* Date
* Year
* Top
* Case
* Count
* Order By
* Group By
* Convert
* Avg
* Null
* Not

## 20 | Categories and Total Products in Each Category 

We want to see the total number of products in each category and sort the results by the total number of products in descending order

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT CategoryName, TotalProducts = COUNT(*) 
FROM Products JOIN Categories ON Products.CategoryID = Categories.CategoryID 
GROUP BY CategoryName 
ORDER BY TotalProducts DESC
```

## 21 | Total Customers per Country / City

Show the total number of customers per country and city.

This table output 25 / 69 rows. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT Country, City, TotalCustomers = Count(*) 
FROM Customers 
GROUP BY Country, City 
ORDER BY TotalCustomers DESC
```

## 22 | Products that Need Reordering 

We want to find the products in our inventory that should be reordered. We can use the fields UnitsInStock and ReorderLEvel where UnitsInStock < ReorderLevel. Sort the results by Product ID

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT ProductID, ProductName, UnitsInStock, ReorderLevel 
FROM Products 
WHERE UnitsInStock <= ReorderLevel 
ORDER BY ProductID
```

## 23 | Products that Need Reordering, Continued 

Now we need to incorporate the fields UnitsInStock, UnitsOnOrder, ReorderLevel, Discontinued into our calculation. 

We can define products that need reordering with the following: 

* UnitsInStock + UnitsOnOrder are less than or equal to ReorderLevel
* The discontinued flag is false (0)

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT ProductID, ProductName, UnitsInStock, UnitsOnOrder, ReorderLevel, Discontinued 
FROM Products 
WHERE NOT (UnitsInStock + UnitsOnOrder) > ReorderLevel 
AND Discontinued = 0
```

## 24 | Customer List by Region 

A salesperson from Northwind is going on a business trip to visit customers, and would like to see a list of all customers, sorted by region, alphabetically. 

However, he wants the customers with no region (null region) to be at the end instead of at the top. Within the same region, companies should be sorted by customer id.

This table outputs 25 / 96 rows. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT CustomerID, CompanyName, Region
FROM Customers
ORDER BY
CASE
  WHEN Region IS NULL THEN 1
  ELSE 0
END, Region
ASC
```

## 25 | High Freight Charges 

Some of the countries shipped to have very high freight charges. Return the three ship countries with the highest average freight overall, in descending order by average freight. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT TOP 3 ShipCountry, AverageFreight = AVG(Freight) 
FROM Orders 
GROUP BY ShipCountry 
ORDER BY AverageFreight DESC
```

## 26 | High Freight Charges for 2015 

Lets return the highest average freight charges from 2015 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT TOP 3 ShipCountry, AverageFreight = AVG(Freight) 
FROM Orders 
WHERE YEAR(CONVERT(DATE, OrderDate)) = 2015 
GROUP BY ShipCountry 
ORDER BY AverageFreight DESC
```

## 27 | High Freight Charges Last Year

We want to see the top three ship countries with highest average freight charges again, but this time we want to use the last 12 months of order data. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT TOP 3 ShipCountry, AverageFreight = AVG(Freight) 
FROM Orders 
WHERE OrderDate BETWEEN DATEADD(yy, -1, (SELECT MAX(OrderDate) FROM Orders)) 
AND (SELECT MAX(OrderDate) FROM Orders) 
GROUP BY ShipCountry 
ORDER BY AverageFreight DESC
```

## 28 | Employee / Order Detail Report 

For inventory, we need to show employee and order detail information for all orders. Sort by Order ID and Product ID.

The table below shows 25 / 2155 rows.

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT Employees.EmployeeID, LastName, Orders.OrderID, Products.ProductName, OrderDetails.Quantity FROM Employees 
JOIN Orders ON Employees.EmployeeID = Orders.EmployeeID
JOIN OrderDetails ON Orders.OrderID = OrderDetails.OrderID
JOIN Products ON OrderDetails.ProductID = Products.ProductID
ORDER BY Orders.OrderID, Products.ProductID
```

## 29 | Customers with No Orders 

We want to return customers who have never placed an order. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT Customers_Customer_ID = Customers.CustomerID, Orders_Customer_ID = Orders.CustomerID 
FROM Customers 
LEFT JOIN Orders ON Orders.CustomerID = Customers.CustomerID
WHERE Orders.CustomerID is NULL
```

## 30 | Customers with No Orders for EmployeeID 4

One employee has placed the most orders (EmployeeID 4: Margaret Peacock). However, there are still some customers who have not placed an order with her. We want to show those customers. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT CustomerID 
FROM Customers 
WHERE NOT EXISTS 
  (SELECT CustomerID 
   FROM Orders
   WHERE Orders.CustomerID = Customers.CustomerID
   AND EmployeeID = 4)
```

# Advanced Problems 

Techniques Used: 

Functions Used: 

## 31 | High Value Customers 

We want to find the customers who have made at least 1 order with a total value (not including discount) of $10,000 or more. We are only considering purchases made in 2016

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT Customers.CustomerID, Customers.CompanyName, Orders.OrderID, TotalPrice = SUM(OrderDetails.UnitPrice * OrderDetails.Quantity)
FROM Customers
JOIN Orders ON Customers.CustomerID = Orders.CustomerID 
JOIN OrderDetails ON Orders.OrderID = OrderDetails.OrderID
WHERE YEAR(CONVERT(DATE, OrderDate)) = 2016
GROUP BY Customers.CustomerID, Customers.CompanyName, Orders.OrderID
HAVING SUM(OrderDetails.UnitPrice * OrderDetails.Quantity) > 10000
ORDER BY TotalPrice DESC
```

## 32 | High Value Customers - Total Orders 

We want to redefine high value customers as those who have orders totalling $15,000 or more in 2016. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT Customers.CustomerID, Customers.CompanyName, TotalPrice = SUM(OrderDetails.UnitPrice * OrderDetails.Quantity)
FROM Customers
JOIN Orders ON Customers.CustomerID = Orders.CustomerID 
JOIN OrderDetails ON Orders.OrderID = OrderDetails.OrderID
WHERE YEAR(CONVERT(DATE, OrderDate)) = 2016
GROUP BY Customers.CustomerID, Customers.CompanyName
HAVING SUM(OrderDetails.UnitPrice * OrderDetails.Quantity) > 15000
ORDER BY TotalPrice DESC
```

## 33 | High Value Customers with Discount

We wish to calcuate high value customers and order by the total amount including discount. 

This table shows 25/81 rows. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT Customers.CustomerID, Customers.CompanyName, TotalsWithoutDiscount = SUM(OrderDetails.UnitPrice * OrderDetails.Quantity), TotalsWithDiscount = ROUND(SUM(OrderDetails.UnitPrice * OrderDetails.Quantity * (1 - OrderDetails.Discount)), 2)
FROM Customers
JOIN Orders ON Customers.CustomerID = Orders.CustomerID 
JOIN OrderDetails ON Orders.OrderID = OrderDetails.OrderID
WHERE YEAR(CONVERT(DATE, OrderDate)) = 2016
GROUP BY Customers.CustomerID, Customers.CompanyName
ORDER BY TotalsWithDiscount DESC
```

## 34 | Month End Orders 

At the end of the month, sales people may try to get more orders to meet a quota. Show all the orders made on the last day of the month, and order it by the employee ID and order ID.

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT EmployeeID, OrderID, OrderDate
FROM Orders
WHERE OrderDate = EOMONTH(OrderDate)
ORDER BY EmployeeID, OrderID
```

## 36 | Orders with Many Line Items

The mobile app developers are testing an app that customers will use to show orders. In order to make sure that even the largest orders show up correctly on the app, they'd like some samples of orders that have lots of individual line items. Show the top 10 orders with the most line items, in order of total line items. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT TOP 10 Orders.OrderID, TotalOrderDetails = COUNT(*)
FROM Orders
JOIN OrderDetails ON Orders.OrderID = OrderDetails.OrderID
GROUP BY Orders.OrderID
ORDER BY COUNT(*) DESC
```

## 37 | Orders - Random Assortment 

The app developers would like to get a random assortment of orders for beta testing on their app. Show a random set of 2% of all orders. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}

```

