---
title: "SQL Practicum"
author: "Michael Rose"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts=list(width.cutoff=60),tidy=TRUE)
options(width = 60)
library(odbc)
library(DBI)
```


```{r}
# rmarkdown::render("sql_practicum.Rmd", params = "ask") 
sql_con <- dbConnect(odbc(),
                 Driver = "ODBC Driver 17 for SQL Server",
                 Server = "localhost",
                 Database = "Northwind_SPP",
                 UID = "SA",
                 PWD = rstudioapi::askForPassword("Database password"),
                 Port = 1433)
```

# Introduction 

This is an in-depth analysis of a production size database using the **MS SQL Server** dialect of SQL. 

The goal is to complete a series of queries that make use of a wide range of the functionality built within sql. 

For the sake of readability, I have limited the output to the top 25 rows. If a query has more than 25 rows, it will be noted above the output.

If you wish to see the full output, in the code there is code chunk options. Remove max.print = 25 for full output.

# Overview

### Why was this project chosen? 

SQL is a very extensible language. Built off a few primitives (select, from, where), the combination of different functions provides a rich toolset for exploring databases. I wanted to build familiarity with some of the features available through practice. 

### What does your project do / how does it work? 

The following project consists of 50 sql queries, ranging in difficulty from simple and moving on top advanced. Each section builds on knowledge built in the last. 

### What tools did you use? 

I used Rmarkdown for the pdf formatting, R language for connecting to the database, microsoft sql server 2017 for setting a database on my localhost, azure data studio for managing permissions for connections, and odbc for connection drivers. The queries themselves are in MS SQL. The database was an extension of the Northwind database modified to contain more tables. The modification was created by **Sylvia Moestl Vasilik**. 

### What did you learn? 

I learned quite a bit of SQL. Thanks to these challenging exercises, I feel much more comfortable with the language and its functionality. I also got to explore the effects of a variety of functions and methods for querying data.

### What is the format of the data?

The data is stored in a MS SQL Server. It contains a variety of tables which are shown in an upcoming section. 

### What other things would you have liked to do with this project?

I would have liked to make some models, but unfortunately this data does not contain any fields worth predicting on. 

# SQL Analysis

## 0.1 | Tables

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT TABLE_NAME
FROM INFORMATION_SCHEMA.TABLES 
WHERE TABLE_TYPE = 'BASE TABLE'
```

## 0.2 | Table Information

```{sql, connection = sql_con, output_var = 'ex1', max.print = NA}
SELECT TABLE_NAME,
       COLUMN_NAME,
       DATA_TYPE
FROM   INFORMATION_SCHEMA.COLUMNS
ORDER BY TABLE_NAME
```

## Categories 

```{sql, connection = sql_con, output_var = 'ex1', max.print = NA}
SELECT TOP 3 * 
FROM Categories
```

## Customers 

```{sql, connection = sql_con, output_var = 'ex1', max.print = NA}
SELECT TOP 3 * 
FROM Customers
```

## Employees

```{sql, connection = sql_con, output_var = 'ex1', max.print = NA}
SELECT TOP 3 * 
FROM Employees
```

## Order Details 

```{sql, connection = sql_con, output_var = 'ex1', max.print = NA}
SELECT TOP 3 * 
FROM OrderDetails
```

## Orders 

```{sql, connection = sql_con, output_var = 'ex1', max.print = NA}
SELECT TOP 3 * 
FROM Orders
```

## Products

```{sql, connection = sql_con, output_var = 'ex1', max.print = NA}
SELECT TOP 3 * 
FROM Products
```

## Shippers 

```{sql, connection = sql_con, output_var = 'ex1', max.print = NA}
SELECT TOP 3 * 
FROM Shippers
```

## Suppliers

```{sql, connection = sql_con, output_var = 'ex1', max.print = NA}
SELECT TOP 3 * 
FROM Suppliers
```


# Easy Queries

Functions Used: 

* Like
    + Determins whether a specific character string matches a specified pattern. 
* Wildcard %
    + Matches any single character within the specified range or set that is specified between brackets. 
* In
    + Determines whether a specified value matches any value in a subquery or list
* Date
    + coerces to date object 
* Order By
    + sorts data returned by a query
* Group By
    + A select statement clause that divides the query result into groups of rows, usually for the purpose of performing one or more aggregations on the group
* Convert
    + coerces to given type
* Concat
    + adds two strings together
* Count
    + returns the number of items in a group
* Min
    + returns the minimum value
* Distinct
    + returns unique values
* Join
    + joins two tables together on a specific column

## 1 | Which Shippers Do We Have?

Return all the fields from all the shippers.

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT * 
FROM Shippers
```

## 2 | Specific Fields from Categories 

Get the categoryname and description from categories. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT CategoryName, Description 
FROM Categories
```

## 3 | Sales Representatives 

Get first name, last name, hire date of all employees with the title of sales representative 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT FirstName, LastName, HireDate 
FROM Employees 
WHERE Title = 'Sales Representative'
```

## 4 | Sales Representatives in the United States 

Constrain to employees in the USA 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT FirstName, LastName, HireDate 
FROM Employees 
WHERE Title = 'Sales Representative' 
AND Country = 'USA'
```

## 5 | Orders Placed by Specific EmployeeID 

Show all orders places by a specific employee. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT OrderID, OrderDate 
FROM Orders 
WHERE EmployeeID = 5
```

## 6 | Suppliers and Contact Titles 

In the suppliers table, we want the supplier ID, contact name and contact title for those suppliers whose contact title is not marketing manager 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT SupplierID, ContactName, ContactTitle 
FROM Suppliers 
WHERE NOT ContactTitle = 'Marketing Manager'
```

## 7 | Products with "Queso" in Product Name

We want the product ID and product name for those products where the product name includes the string Queso 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT ProductID, ProductName 
FROM Products 
WHERE ProductName 
LIKE '%queso%'
```

## 8 | Orders Shipping to France or Belgium 

We want the order id, customer id, and ship country for the orders where the ship country is either France or Belgium.
The following table shows 25 / 96 rows.

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT OrderID, CustomerID, ShipCountry 
FROM Orders 
WHERE ShipCountry 
IN ('France', 'Belgium')
```

## 9 | Orders Shipping to any Country in Latin America

We want to get all the orders from any latin american country | We don't have a list of the Latin American countries in a table, so we will make it.

This table outputs 25 / 173 rows. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT OrderID, CustomerID, ShipCountry 
FROM Orders 
WHERE ShipCountry 
IN ('Brazil', 'Mexico', 'Argentina', 'Venezuela')
```

## 10 | Employees in Order of Age 

For all the employees, show first, last, title and birth date in order of birth date with oldest employees first

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT FirstName, LastName, Title, BirthDate 
FROM Employees 
ORDER BY BirthDate ASC
```

## 11 | Showing Only the Date with a DateTime Field 

In the above query, suppose we only want the date portion of the birth date field. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT FirstName, LastName, Title, BDate = convert(DATE, BirthDate) 
FROM Employees 
ORDER BY BDate ASC
```

## 12 | Employees Full Name
 
show first, last and firstlast names

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT FirstName, LastName, FullName = CONCAT(FirstName, ' ', LastName) 
FROM Employees
```

## 13 | Order Details Amount per Line Item 

In the order details table we have the fields unit price and quantity. We want a new field, total price that multiplies these two together. We also want order id, product id, unit price, and quantity ordered by order id and product id.

This table outputs 25 / 2155 rows. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT OrderID, ProductID, UnitPrice, Quantity, TotalPrice = UnitPrice * Quantity 
FROM OrderDetails 
ORDER BY OrderID, ProductID
```

## 14 | How Many Customers? 

How many customers are there in the customers table? 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT TotalCustomers = COUNT(*) 
FROM Customers
```

## 15 | Date of First Order

Return the date of the first order 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT FirstOrder = MIN(OrderDate) 
FROM Orders
```

## 16 | Countries with Customers 

We want a list of countries where Northwind has customers 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT DISTINCT(Country) 
FROM Customers
```

## 17 | Contact Titles for Customers 

Return a list of all the different values in the customers table for contact titles. Also include a count for each contact title. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT ContactTitle, TotalContactTitle = COUNT(*) 
FROM Customers 
GROUP BY ContactTitle 
ORDER BY TotalContactTitle DESC
```

## 18 | Products with Associated Supplier Names 

For each product, we want the associated supplier. Show the product id, product name, and company name of the supplier and sort by product id. 

This table outputs 25 / 77 rows. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT ProductID, ProductName, Supplier = CompanyName 
FROM Products 
JOIN Suppliers ON Products.SupplierID = Suppliers.SupplierID
```

## 19 | Orders and the Shipper Used 

We'd like to show a list of the orders that were made, including the shipper that was used. We want order id, order date (date only), company name of the shipper, and to sort by order id with order id < 10270. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT OrderID, OrderDate = CONVERT(DATE, ORDERDATE), Shipper = CompanyName 
FROM Orders JOIN Shippers ON Orders.ShipVia = Shippers.ShipperID 
WHERE OrderID < 10270 ORDER BY OrderID
```

# Intermediate Problems 

Functions Used: 

* Exists 
    + specifies a subquery to test for the existence of rows
* DateAdd
    + Adds a specified number value to a specified datepart of an input date value, and then returns that modified value
* Nested Select
    + allows for a query as an operated on dataframe
* Multiple Join
    + joins multiple tables on specified column values
* Date
    + coerces to date
* Year
    + pulls the year value from a date object
* Top
    + returns the top n rows
* Case
    + allows for a control sequence on data
* Count
    + returns number of items from a group
* Order By
    + sorts the data
* Group By
    + A select statement clause that divides the query result into groups of rows, usually for the purpose of performing one or more aggregations on the group
* Convert
    + coerces to a given datatype
* Avg
    + returns the average of a column of numbers
* Not
    + returns the negation of the statement

## 20 | Categories and Total Products in Each Category 

We want to see the total number of products in each category and sort the results by the total number of products in descending order

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT CategoryName, TotalProducts = COUNT(*) 
FROM Products JOIN Categories ON Products.CategoryID = Categories.CategoryID 
GROUP BY CategoryName 
ORDER BY TotalProducts DESC
```

## 21 | Total Customers per Country / City

Show the total number of customers per country and city.

This table output 25 / 69 rows. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT Country, City, TotalCustomers = Count(*) 
FROM Customers 
GROUP BY Country, City 
ORDER BY TotalCustomers DESC
```

## 22 | Products that Need Reordering 

We want to find the products in our inventory that should be reordered. We can use the fields UnitsInStock and ReorderLEvel where UnitsInStock < ReorderLevel. Sort the results by Product ID

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT ProductID, ProductName, UnitsInStock, ReorderLevel 
FROM Products 
WHERE UnitsInStock <= ReorderLevel 
ORDER BY ProductID
```

## 23 | Products that Need Reordering, Continued 

Now we need to incorporate the fields UnitsInStock, UnitsOnOrder, ReorderLevel, Discontinued into our calculation. 

We can define products that need reordering with the following: 

* UnitsInStock + UnitsOnOrder are less than or equal to ReorderLevel
* The discontinued flag is false (0)

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT ProductID, ProductName, UnitsInStock, UnitsOnOrder, ReorderLevel, Discontinued 
FROM Products 
WHERE NOT (UnitsInStock + UnitsOnOrder) > ReorderLevel 
AND Discontinued = 0
```

## 24 | Customer List by Region 

A salesperson from Northwind is going on a business trip to visit customers, and would like to see a list of all customers, sorted by region, alphabetically. 

However, he wants the customers with no region (null region) to be at the end instead of at the top. Within the same region, companies should be sorted by customer id.

This table outputs 25 / 96 rows. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT CustomerID, CompanyName, Region
FROM Customers
ORDER BY
CASE
  WHEN Region IS NULL THEN 1
  ELSE 0
END, Region
ASC
```

## 25 | High Freight Charges 

Some of the countries shipped to have very high freight charges. Return the three ship countries with the highest average freight overall, in descending order by average freight. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT TOP 3 ShipCountry, AverageFreight = AVG(Freight) 
FROM Orders 
GROUP BY ShipCountry 
ORDER BY AverageFreight DESC
```

## 26 | High Freight Charges for 2015 

Lets return the highest average freight charges from 2015 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT TOP 3 ShipCountry, AverageFreight = AVG(Freight) 
FROM Orders 
WHERE YEAR(CONVERT(DATE, OrderDate)) = 2015 
GROUP BY ShipCountry 
ORDER BY AverageFreight DESC
```

## 27 | High Freight Charges Last Year

We want to see the top three ship countries with highest average freight charges again, but this time we want to use the last 12 months of order data. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT TOP 3 ShipCountry, AverageFreight = AVG(Freight) 
FROM Orders 
WHERE OrderDate BETWEEN DATEADD(yy, -1, (SELECT MAX(OrderDate) FROM Orders)) 
AND (SELECT MAX(OrderDate) FROM Orders) 
GROUP BY ShipCountry 
ORDER BY AverageFreight DESC
```

## 28 | Employee / Order Detail Report 

For inventory, we need to show employee and order detail information for all orders. Sort by Order ID and Product ID.

The table below shows 25 / 2155 rows.

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT Employees.EmployeeID, LastName, Orders.OrderID, Products.ProductName, OrderDetails.Quantity 
FROM Employees 
JOIN Orders ON Employees.EmployeeID = Orders.EmployeeID
JOIN OrderDetails ON Orders.OrderID = OrderDetails.OrderID
JOIN Products ON OrderDetails.ProductID = Products.ProductID
ORDER BY Orders.OrderID, Products.ProductID
```

## 29 | Customers with No Orders 

We want to return customers who have never placed an order. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT Customers_Customer_ID = Customers.CustomerID, Orders_Customer_ID = Orders.CustomerID 
FROM Customers 
LEFT JOIN Orders ON Orders.CustomerID = Customers.CustomerID
WHERE Orders.CustomerID is NULL
```

## 30 | Customers with No Orders for EmployeeID 4

One employee has placed the most orders (EmployeeID 4: Margaret Peacock). However, there are still some customers who have not placed an order with her. We want to show those customers. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT CustomerID 
FROM Customers 
WHERE NOT EXISTS 
  (SELECT CustomerID 
   FROM Orders
   WHERE Orders.CustomerID = Customers.CustomerID
   AND EmployeeID = 4)
```

# Advanced Problems 

Functions Used: 

* sum
    + returns a sum of a column of numeric values

* multiple join
    + joins multiple tables by column 

* convert
    + coerces to a given datatype

* date
    + coerces to a date object

* group by
    + A select statement clause that divides the query result into groups of rows, usually for the purpose of performing one or more aggregations on the group

* order by
    + sorts the given data by specified values

* round
    + rounds to nearest given integer expansion, e.g. whole number, 0.1, 0.01, 0.001, ...

* eomonth
    + returns the last day of each months value

* top
    + returns the top n rows

* newid
    + creates a unique identifier for the assigned value

* percent
    + returns a percentage of the data as opposed to a fixed integer

* count
    + returns a sum of a specified data set

* subquery
    + a query within a query

* common table expressions
    + a query created specifically for aid in another query

* distinct 
    + returns unique values

* isnull
    + if a value is null, coerces null to a given number

* case
    + allows for control flow on a given computation

* union
    + a set operation that returns all unique values of multiple given queries

* full outer join
    + a join which keeps the values of both joined tables, regardless of whether the items are in both tables or not

* row_number
    + adds a row number identifier

* window partition
    + partitions the data into segments and allows for a set of functions to be performed over the partitions

* table alias
    + allows for multiple named instances of a single database to be modified independently and used in a query


## 31 | High Value Customers 

We want to find the customers who have made at least 1 order with a total value (not including discount) of $10,000 or more. We are only considering purchases made in 2016

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT Customers.CustomerID, Customers.CompanyName, Orders.OrderID, TotalPrice = SUM(OrderDetails.UnitPrice * OrderDetails.Quantity)
FROM Customers
JOIN Orders ON Customers.CustomerID = Orders.CustomerID 
JOIN OrderDetails ON Orders.OrderID = OrderDetails.OrderID
WHERE YEAR(CONVERT(DATE, OrderDate)) = 2016
GROUP BY Customers.CustomerID, Customers.CompanyName, Orders.OrderID
HAVING SUM(OrderDetails.UnitPrice * OrderDetails.Quantity) > 10000
ORDER BY TotalPrice DESC
```

## 32 | High Value Customers - Total Orders 

We want to redefine high value customers as those who have orders totalling $15,000 or more in 2016. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT Customers.CustomerID, Customers.CompanyName, TotalPrice = SUM(OrderDetails.UnitPrice * OrderDetails.Quantity)
FROM Customers
JOIN Orders ON Customers.CustomerID = Orders.CustomerID 
JOIN OrderDetails ON Orders.OrderID = OrderDetails.OrderID
WHERE YEAR(CONVERT(DATE, OrderDate)) = 2016
GROUP BY Customers.CustomerID, Customers.CompanyName
HAVING SUM(OrderDetails.UnitPrice * OrderDetails.Quantity) > 15000
ORDER BY TotalPrice DESC
```

## 33 | High Value Customers with Discount

We wish to calcuate high value customers and order by the total amount including discount. 

This table shows 25/81 rows. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT Customers.CustomerID, Customers.CompanyName, 
       TotalsWithoutDiscount = SUM(OrderDetails.UnitPrice * OrderDetails.Quantity), 
       TotalsWithDiscount = ROUND(SUM(OrderDetails.UnitPrice * OrderDetails.Quantity * (1 - OrderDetails.Discount)), 2)
FROM Customers
JOIN Orders ON Customers.CustomerID = Orders.CustomerID 
JOIN OrderDetails ON Orders.OrderID = OrderDetails.OrderID
WHERE YEAR(CONVERT(DATE, OrderDate)) = 2016
GROUP BY Customers.CustomerID, Customers.CompanyName
ORDER BY TotalsWithDiscount DESC
```

## 34 | Month End Orders 

At the end of the month, sales people may try to get more orders to meet a quota. Show all the orders made on the last day of the month, and order it by the employee ID and order ID.

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT EmployeeID, OrderID, OrderDate
FROM Orders
WHERE OrderDate = EOMONTH(OrderDate)
ORDER BY EmployeeID, OrderID
```

## 35 | Orders with Many Line Items

The mobile app developers are testing an app that customers will use to show orders. In order to make sure that even the largest orders show up correctly on the app, they'd like some samples of orders that have lots of individual line items. Show the top 10 orders with the most line items, in order of total line items. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT TOP 10 Orders.OrderID, TotalOrderDetails = COUNT(*)
FROM Orders
JOIN OrderDetails ON Orders.OrderID = OrderDetails.OrderID
GROUP BY Orders.OrderID
ORDER BY COUNT(*) DESC
```

## 36 | Orders - Random Assortment 

The app developers would like to get a random assortment of orders for beta testing on their app. Show a random set of 2% of all orders. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT TOP 2 PERCENT OrderID
FROM Orders
ORDER BY NEWID()
```

## 37 | Orders - Accidental Double Entry

Suppose someone came to you claiming they accidentally entered a line item twice on an order, each time with a different product ID, but the same quantity. They remember the quantity was 60 or more. Show all the order IDs that match this, in order of order ID.

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT OrderID, Quantity
FROM OrderDetails
WHERE Quantity >= 60
GROUP BY OrderID, Quantity
HAVING COUNT(OrderID) > 1
ORDER BY OrderID
```

## 38 | Orders - Accidental Double Entry with Details 

We want to show the details of the order for orders that match the criteria of the above query. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
;WITH Acc_Doub_Entry AS(
  SELECT OrderID, Quantity
  FROM OrderDetails
  WHERE Quantity >= 60
  GROUP BY OrderID, Quantity
  HAVING COUNT(OrderID) > 1)
SELECT OrderID, ProductID, UnitPrice, Quantity, Discount
FROM OrderDetails
WHERE OrderID IN (SELECT OrderID FROM Acc_Doub_Entry)
ORDER BY OrderID, Quantity
```

## 39 | Orders - Accidental Double Entry Details Derived Table 

How would we get the same results as above without a common table expression?

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT OrderDetails.OrderID, ProductID, UnitPrice, Quantity, Discount
FROM OrderDetails
JOIN (SELECT DISTINCT(OrderID)
FROM OrderDetails
WHERE Quantity >= 60
GROUP BY OrderID, Quantity
HAVING COUNT(OrderID) > 1) PotentialProblemOrders
ON PotentialProblemOrders.OrderID = OrderDetails.OrderID
ORDER BY OrderID, ProductID
```

## 40 | Late Orders 

Find all the orders that are arriving late. Place them in order by the longest delays to the shortest delays.

The table shows 25/39 rows.

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT OrderID, OrderDate = CONVERT(DATE, OrderDate), RequiredDate, ShippedDate 
FROM Orders
WHERE CONVERT(DATE, ShippedDate) >= CONVERT(DATE, RequiredDate)
ORDER BY DATEDIFF(dd, CONVERT(DATE, ShippedDate), CONVERT(DATE, RequiredDate)) ASC
```

## 41 | Late Orders by Employees

Some salespeople have more orders arriving late than others. Which sales people have the most orders arriving late?

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT Employees.EmployeeID, Employees.LastName, TotalLateOrders = COUNT(*)
FROM Orders
JOIN Employees ON Orders.EmployeeID = Employees.EmployeeID
WHERE CONVERT(DATE, ShippedDate) >= CONVERT(DATE, RequiredDate)
GROUP BY Employees.EmployeeID, Employees.LastName
ORDER BY TotalLateOrders DESC
```

## 42 | Late Orders vs. Total Orders 

We want to compare the number of late orders per salesperson with the number of total orders per salesperson. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
;WITH Tot_Late_Orders AS(
  SELECT Employees.EmployeeID, Employees.LastName, TotalLateOrders = COUNT(*)
  FROM Orders
  JOIN Employees ON Orders.EmployeeID = Employees.EmployeeID
  WHERE CONVERT(DATE, ShippedDate) >= CONVERT(DATE, RequiredDate)
  GROUP BY Employees.EmployeeID, Employees.LastName
  ),

Overall_Orders AS(
  SELECT Orders.EmployeeID, Employees.LastName, TotalOrders = COUNT(*)
  FROM Orders
  JOIN Employees ON Orders.EmployeeID = Employees.EmployeeID
  GROUP BY Orders.EmployeeID, Employees.LastName
)

SELECT Overall_Orders.EmployeeID, Overall_Orders.LastName, AllOrders = TotalOrders, 
       LateOrders = ISNULL(TotalLateOrders, 0), 
       Prop = ROUND(((ISNULL(TotalLateOrders, 0) * 1.0) / Overall_Orders.TotalOrders), 2)
FROM Tot_Late_Orders
RIGHT JOIN Overall_Orders ON Tot_Late_Orders.EmployeeID = Overall_Orders.EmployeeID
ORDER BY Prop DESC
```

## 43 | Customer Grouping

The VP of Sales would like us to create customer groups depending on how much they ordered in 2016. The groupings would look like the following: 

* 0 - 1000
* 1001 - 5000
* 5001 - 10000
* 10000+

The table below shows 25/81 rows.

```{sql, connection = sql_con, output_var = 'ex1'}
SELECT Customers.CustomerID, Customers.CompanyName, 
       TotalPrice = SUM(OrderDetails.UnitPrice * OrderDetails.Quantity), 
       CustomerGroup = CASE
       WHEN SUM(OrderDetails.UnitPrice * OrderDetails.Quantity) <= 1000 THEN 'Low'
       WHEN SUM(OrderDetails.UnitPrice * OrderDetails.Quantity) > 1000 
       AND SUM(OrderDetails.UnitPrice * OrderDetails.Quantity) <= 5000 THEN 'Medium'
       WHEN SUM(OrderDetails.UnitPrice * OrderDetails.Quantity) > 5000 
       AND SUM(OrderDetails.UnitPrice * OrderDetails.Quantity) <= 10000 THEN 'High'
       ELSE 'Very High'
END
FROM Customers
JOIN Orders ON Customers.CustomerID = Orders.CustomerID 
JOIN OrderDetails ON Orders.OrderID = OrderDetails.OrderID
WHERE YEAR(CONVERT(DATE, OrderDate)) = 2016
GROUP BY Customers.CustomerID, Customers.CompanyName
ORDER BY Customers.CustomerID ASC
```

## 44 | Customer Grouping with Percentage

We would like to show the breakdown of each group and their respective percentages of the whole catalog of business.

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
;WITH Orders_2016 AS(
  SELECT Customers.CustomerID, Customers.CompanyName, TotalOrderAmount = SUM(Quantity * UnitPrice)
  FROM Customers 
  JOIN Orders ON Orders.CustomerID = Customers.CustomerID
  JOIN OrderDetails ON Orders.OrderID = OrderDetails.OrderID
  WHERE YEAR(CONVERT(DATE, OrderDate)) = 2016
  GROUP BY Customers.CustomerID, Customers.CompanyName
),
Cust_Groups AS(
  SELECT CustomerID, CompanyName, TotalOrderAmount, CustomerGroup = CASE
    WHEN TotalOrderAmount <= 1000 THEN 'Low'
    WHEN TotalOrderAmount > 1000 AND TotalOrderAmount <= 5000 THEN 'Medium'
    WHEN TotalOrderAmount > 5000 AND TotalOrderAmount <= 10000 THEN 'High'
    ELSE 'Very High'
  END
  FROM Orders_2016)
  
SELECT CustomerGroup, TotalInGroup = COUNT(*), 
       PercentageInGroup = COUNT(*) * 1.0 / (SELECT COUNT(*) FROM Cust_Groups)
FROM Cust_Groups
GROUP BY CustomerGroup
ORDER BY TotalInGroup DESC
```

## 45 | Customer Grouping - Flexible 

The sales team would like an overview of the customer groupings, and would like them in a way in which they don't need to edit much sql in order to change the boundaries. 

In this table we will use the CustomerGroupThresholds table. The table output is 25/81 rows.

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
;WITH Orders_2016 AS (
  SELECT Customers.CustomerID, Customers.CompanyName, TotalOrderAmount = SUM(Quantity * UnitPrice)
  FROM Customers
  JOIN Orders ON Orders.CustomerID = Customers.CustomerID
  JOIN OrderDetails ON Orders.OrderID = OrderDetails.OrderID
  WHERE YEAR(CONVERT(DATE, OrderDate)) = 2016
  GROUP BY Customers.CustomerID, Customers.CompanyName
)
SELECT CustomerID, CompanyName, TotalOrderAmount, CustomerGroupName 
FROM Orders_2016
JOIN CustomerGroupThresholds ON Orders_2016.TotalOrderAmount
BETWEEN CustomerGroupThresholds.RangeBottom AND CustomerGroupThresholds.RangeTop
ORDER BY CustomerID
```

## 46 | Countries with Suppliers or Customers 

Return a list of all countries where suppliers or customers are based.

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT Distinct(Country) 
FROM Customers
UNION 
SELECT DISTINCT(Country)
FROM Suppliers
ORDER BY Country
```

## 47 | Countries with Suppliers or Customers V2

With More Details

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
;WITH SupplierCountries AS(
SELECT DISTINCT(Country) FROM Suppliers
),
CustomerCountries AS(
SELECT DISTINCT(Country) FROM Customers
)
SELECT SupplierCountry = SupplierCountries.Country, CustomerCountry = CustomerCountries.Country
FROM SupplierCountries
FULL OUTER JOIN CustomerCountries ON SupplierCountries.Country = CustomerCountries.Country
```

## 48 | Countries with Suppliers or Customers V3 

Return country name, total suppliers, and total customers. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
;WITH CountriesList AS(
  SELECT Distinct(Country) 
  FROM Customers
  UNION
  SELECT DISTINCT(Country)
  FROM Suppliers
), 
SupplierCountries AS(
SELECT Country, NumCountry = COUNT(*) FROM Suppliers
GROUP BY Country
),
CustomerCountries AS(
SELECT Country, NumCountry = COUNT(*) FROM Customers
GROUP BY Country
)
SELECT CountriesList.Country, 
       TotalSuppliers = ISNULL(SupplierCountries.NumCountry, 0), 
       TotalCustomers = ISNULL(CustomerCountries.NumCountry, 0)
FROM CountriesList
FULL OUTER JOIN SupplierCountries ON CountriesList.Country = SupplierCountries.Country
FULL OUTER JOIN CustomerCountries ON CountriesList.Country = CustomerCountries.Country
ORDER BY Country
```

## 49 | First Order in Each Country 

Suppose we wish to know the first order in each country, ordered by order id. 

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
;WITH OrdersByCountry AS(
  SELECT ShipCountry, CustomerID, OrderID, OrderDate = CONVERT(DATE, OrderDate), 
  RowNumberPerCountry = Row_Number() OVER(PARTITION BY ShipCountry ORDER BY ShipCountry, OrderID)
  FROM Orders
)
SELECT ShipCountry, CustomerID, OrderID, OrderDate
FROM OrdersByCountry
WHERE RowNumberPerCountry = 1
ORDER BY ShipCountry
```

## 50 | Customers with Multiple Orders in 5 Day Period

There are customers for whom freight is a major expense when ordering. By batching up their orders, and making one large order as opposed to multiple smaller orders in a short period of time, they can reduce their freight costs significantly. 

Show the customers who have made more than 1 order in a 5 day period. The table shows 25 / 71 rows.

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
SELECT InitialOrder.CustomerID, 
       InitialOrderID = InitialOrder.OrderID,
       InitialOrderDate = CONVERT(DATE, InitialOrder.OrderDate),
       NextOrderID = NextOrder.OrderID, 
       NextOrderDate = CONVERT(DATE, NextOrder.OrderDate),
       DaysBetweenOrders = DATEDIFF(dd, InitialOrder.OrderDate, NextOrder.OrderDate)
FROM Orders InitialOrder
JOIN Orders NextOrder ON InitialOrder.CustomerID = NextOrder.CustomerID
WHERE InitialOrder.OrderID < NextOrder.OrderID
AND DATEDIFF(dd, InitialOrder.OrderDate, NextOrder.OrderDate) <= 5
ORDER BY InitialOrder.CustomerID, InitialOrder.OrderID
```

Alternatively, we could do it with a window function. The table below shows 25 / 69 rows.

```{sql, connection = sql_con, output_var = 'ex1', max.print = 25}
;WITH NextOrderDate AS(
  SELECT CustomerID, OrderDate = CONVERT(DATE, OrderDate),
  NextOrderDate = CONVERT(DATE, LEAD(OrderDate, 1) 
  OVER(PARTITION BY CustomerID ORDER BY CustomerID, OrderDate))
  FROM Orders
)
SELECT CustomerID, OrderDate, NextOrderDate, DaysBetweenOrders = DATEDIFF(dd, OrderDate, NextOrderDate)
FROM NextOrderDate
WHERE DATEDIFF(dd, OrderDate, NextOrderDate) <= 5
```
